---
title: "Working with data and data.table"
author: "Megan Stodel"
date: "27 May 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's look at the hawks data base we used for the `ggplot2` workshop. With the `data.table` library, we can use `fread()` to read in the csv, which will be quick and give us a data.table object. 

```{r}
library(data.table)

hawks <- fread('https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Hawks.csv', 
               select = c('Species', 'Age', 'Sex', 'Wing', 'Weight', 'Tail', 'KeelFat', 'Tarsus', 'Year'))


# View the first five rows
hawks[1:5]
```

### Missing data

There are a few ways to look for missing data. We can use the `is.na()` function when filtering by rows, like this:

```{r}
hawks[is.na(Weight)]
```

This shows us rows where Weight has an NA value. You can do this across all columns with `complete.cases()`, which is from the base stats package, negating it with a `!` to find incomplete cases:

```{r}
hawks[!complete.cases(hawks)]
```

OK, there are a LOT of incomplete cases. Looking at the data it seems likely that KeelFat and Tarsus are full of missing data. 

```{r}
hawks[is.na(KeelFat), .N]
hawks[is.na(Tarsus), .N]
```

We might decide for our analysis we don't want to include those variables if they are that incomplete. There are a couple of ways to get rid of multiple columns. You can either select a list of the columns you do want:

```{r}
# To save this you would need to assign it to an object
hawks[, .(Species, Age, Sex, Wing, Weight, Tail, Year)]
```

Or instead set the columns you want to get rid of to `NULL`:: 

```{R}
hawks[, c("KeelFat", "Tarsus") := NULL]

# No need to reassign
hawks[1:5]
```

Let's review the incomplete cases again.

```{r}
hawks[!complete.cases(hawks)]
```

We could probably get rid of those; there are now only 11 out of 908 observations. 

```{r}
hawks <- hawks[complete.cases(hawks)]
hawks[is.na(Weight)]
```

But the very observant will have noticed that there is also missing data in some of the columns that doesn't seem to be registering. For example, the first row of the dataset isn't in this summary of incomplete cases, even though it's missing information on sex. 

```{r}
hawks[1]
```

That's because the Sex variable isn't an NA value; it appears to just be an empty string. Let's check that by looking at the unique values for Sex. 

```{r}
hawks[, unique(Sex)]
```

As suspected, it's an empty string rather than an NA value. But how many are there? 

```{r}
hawks[, .N, by = Sex]
```

Oh dear - hawks with unknown sex make up most of the dataset, so we can't really just remove them. For now, we'll make it a bit more explicit the data is unknown. Note that we are using `:=` so we don't have to reassign the data.table object. 

```{r}
hawks[Sex == "", Sex := "Unknown"]
hawks[1:5]
```

We should check the other character columns. 

```{r}
hawks[, .N, by = Species]
hawks[, .N, by = Age]
```

Great, both Species and Age are complete columns. 

### Changing category values

We saw above how we could change a value by filtering it first. Let's do that for the other sex values.

```{r}
hawks[Sex == "F", Sex := "Female"]
hawks[Sex == "M", Sex := "Male"]
```

We can then change this to a factor variable, to ensure only these values can be used. 

```{r}
hawks[, Sex := factor(Sex, levels = c("Female", "Male", "Unknown"))]
levels(hawks[, Sex])
```

And the same for Species, but this time redefining the values after they've become factors so you see both ways. 

```{r}
hawks[, Species := factor(Species, levels = c("CH", "RT", "SS"))]
levels(hawks$Species) <- c("Cooper's", "Ring-tailed", "Sharp-shinned")

levels(hawks[, Species])
```

### Exploring the data

So far we've been looking at the data for issues that we want to clean up, but you can use `data.table` to answer questions you have about the data too. Here's an example to find out which species had the most adults recorded in the year 2000. 

```{r}
# Filter by year, then count cases while grouping by species and age
age_species_2000 <- hawks[Year == 2000, .(Count = .N), by = .(Species, Age)]
age_species_2000

# Filter to adults
adult_species_2000 <- age_species_2000[Age == "A"]
adult_species_2000

# Order by count
ordered <- adult_species_2000[order(-Count)]
ordered

# Select the value for the species in the first row
most_adults <- ordered[1, Species]
most_adults

```

See how you can pipe this all together if you want:

```{r}
hawks[Year == 2000, .(Count = .N), by = .(Species, Age)][Age == "A"][order(-Count)][1, Species]
```

### Excercises

1. Change the Age variable so it is a factor with the levels Immature and Adult. 
2. Make a new boolean column that indicates if wing length is more than 400.
3. Rename 'Tail' to 'TailLength'.
4. Which year had the highest mean weight of hawk? (use the `mean()` function)
5. Make every column a character class. 


```{r, eval = FALSE, echo = FALSE}
# Change the Age variable so it is a factor with the levels Immature and Adult. 
hawks[, Age := factor(Age, levels = c("I", "A"))]
levels(hawks$Age) <- c("Immature", "Adult")

# Make a new boolean column that indicates if wing length is more than 400.
hawks[, LongWing := Wing > 400]

# Rename 'Tail' to 'TailLength'.
setnames(hawk, "Tail", "TailLength")

# Which year had the highest mean weight of hawk?
hawks[, .(MeanWeight = mean(Weight)), by = Year][order(-MeanWeight)][1, Year]

# Make every column a character class. 
hawks_ch <- hawks[, lapply(.SD, as.character)]


```